---
title: "Lab 3: Probability Models"
subtitle: "BIOL/OCSC 4605 · 7220 --- Statistical Analysis in Biology and Environmental Science"
author: "Your Name"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: flatly
    self_contained: true
params:
  week: 3
  module: "Part 2: Quantifying Uncertainty"
---

<!--
  ============================================================================
  INSTRUCTOR WALKTHROUGH -- BIOL/OCSC 4605 · 7220, Lab 3 "Probability Models"
  ============================================================================
  This file is a live-coding companion: this
  lab coverers the binomial pdf and cdf, likelihood and likelihood ratios, and 
  p-values / critical values for chi-squared, t, F, and the Normal (Z) distributions.
  Section numbers and headings below match the handout's own numbering (Ex. 1-14).
  All computations use base R only (dbinom/pbinom/qbinom, dchisq/pchisq/
  qchisq, dt/pt/qt, df/pf/qf, dnorm/pnorm/qnorm). No packages to install.
  Knit to PDF or HTML from the dropdown next to the Knit button in RStudio.
  ============================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                       fig.width = 5, fig.height = 4, fig.align = "center")
```


## Warm-up: your package as a calculator

*Handout: "For a coin toss, the probability of Heads on the first toss is p = ___. Now calculate the probability of Heads on 2 successive tosses, p² = ___."*

```{r warmup}
p <- 0.5
p          # probability of Heads on one toss
p^2        # probability of Heads on 2 successive (independent) tosses
```

## Exercise 1: independent and dependent events

*Handout: "Now calculate the probability of 4 Heads on 4 tosses."*

```{r ex1-four-heads}
# write the code to answer the question

```

*Handout: card draws without replacement — probability of 2 successive diamonds from a 52-card deck.*

```{r ex1-cards}
# P(diamond on the first draw)
# How many diamonds in total in a card deck?
# Ace, 2 through 10, Jack, Queen, and King (n/total cards)
p_diamond_1 <- NA/NA

# How many diamonds remain after removing one?
diamonds_remaining <- NA

# How many cards remain after removing one?
cards_remaining <- NA

# What is the probability of another diamond?
# This is called a conditional probability
p_diamond_2_given_1 <- diamonds_remaining / cards_remaining  

p_diamond_1
p_diamond_2_given_1

# Probability of 2 successive diamonds = product of the two (dependent) events
p_two_diamonds <- p_diamond_1 * p_diamond_2_given_1
p_two_diamonds
```

**Note:** Exercise 1's first part (coin tosses) multiplies two *independent* probabilities; the card draw multiplies a probability by a *conditional* probability, because removing a card changes the deck. Same multiplication rule, different justification.

## Exercise 2: a single binomial probability (the planarian T-maze)

In 1959 James V. McConnell created The Worm Runner's Digest (W.R.D.) to report his experiments with memory transfer in planarian worms (see Hartry AL et al 1964 Planaria: Memory transfer through cannibalism re-examined Science 146: 274-275).

What is the probability that an untrained planarian worm will correctly guess whether to turn left or right to obtain food in a T-maze on 6 successive trials? I.e., what is the probability of 6 correct guesses by chance alone?

*Handout: "What is the probability of 6 correct guesses by chance alone, out of 6 trials, with p = 0.5?"*
*First by hand, then with `dbinom`.*

```{r ex2-single-binomial}
# By hand, for a run of all successes: f(x) = p^n when x = n
# find p, which is 1 correct guess then p^n
# write your math

0.3^10

# The general binomial pdf, via R's built-in function:
# Find what the function does by running ?dbinom in the console
# x is the number of successes
# size = the number of trials
# prob = probability of success on each trial.
dbinom(x = 6, size = 6, prob = 0.5)
```

*Handout: "Now calculate the probability of a worm making 5 correct choices in 6 trials assuming a success rate of 50%*

*Handout: "Calculate the probability of a worm making 7 correct choices in 7 trials assuming a success rate of 80%*


```{r ex2-more-binomials}
# 5 correct choices in 6 trials
dbinom(x = NA, size = NA, prob = NA)
# 7 correct choices in 7 trials
dbinom(x = NA, size = NA, prob = NA)
```

## Exercise 3: the full binomial probability distribution

What is a PDF?

PDF stands for Probability Density Function. It is an important concept in statistics for understanding probabilities related to continuous random variables. It is a smooth curve that shows how likely different outcomes are within a range of values.

*Handout: build a column `x = 0..6` and a matching column `f(x)`, then plot it.*

*Run the R commands below line by line*


```{r ex3-binomial-distribution}
# Create a vector of possible outcomes (0 to 6 successes)
x <- 0:6
x
# Calculate the binomial probability for each outcome in x
# for n = 6 trials (size) and probability of success p = 0.5 (prob)
f.x <- dbinom(x, size = 6, prob = 0.5)
f.x
# Combine x and f.x column-wise into a matrix to view outcomes alongside their probabilities
cbind(x, f.x)

# Check: the pdf must sum to 1 over all possible outcomes
# Sum all calculated probabilities in f.x to verify they sum to 1.0
sum(f.x)
```

```{r ex3-barplot, fig.cap="Binomial pdf, n=6, p=0.5"}
barplot(f.x, names.arg = x, xlab = "x (successes in 6 trials)",
        ylab = "f(x)", col = "steelblue3",
        main = "Pr{X = x}, n = 6, p = 0.5")
```

## Exercise 4: likelihood and the likelihood ratio in games of chance (roulette, dice, and cards)

*The likely value for the parameter $\pi$ ($\hat{\pi}$)*

*Comapred to the previous experiment, instead of calculating the probability given a fixed parameter we calculate the likelihood of a parameter, given the data*

*Handout: two experiments (9 successes in 10 trials; 7 successes in 10 trials). Calculate the likelihood $\mathcal{L}(\hat{\pi}\,|\,data)$ over a range of candidate $\pi$ values from 0.999 down to 0.50, find the maximum, and compute the likelihood ratio against $\pi = 0.5$.*

```{r ex4-likelihood-setup}
# candidate likelihood with pi = 0.5 (no learning) and pi = 0.999 (highly trained) 
pi_candidates <- c(0.999, 0.95, 0.90, 0.85, 0.80, 0.75, 0.70, 0.65, 0.60, 0.55, 0.50)
# Experiment 1: 9 success over 10 trials
L.exp1 <- dbinom(x = 9, size = 10, prob = pi_candidates)
# Experiment 2: 7 success over 10 trials
L.exp2 <- dbinom(x = 7, size = 10, prob = pi_candidates)

likelihood_table <- data.frame(
  pi = pi_candidates,
  L_Experiment1 = round(L.exp1, 5),
  L_Experiment2 = round(L.exp2, 5)
)
likelihood_table
```

*Maximum Likelihood*

```{r ex4-max-likelihood}
# For the first experiment, what is the maximum likelihood?
# Experiment 1 (9 successes in 10 trials)
max_L1 <- max(L.exp1)
# Find the \hat_pi => the max likelihood
pi_hat_1 <- pi_candidates[which.max(L.exp1)]

L1_at_no_learning <- dbinom(9, 10, 0.5)

# Calculate the likelihood ratio LR = Max($\hat \pi$|data) / L($\pi=0.5$|data)
LR1 <- max_L1 / L1_at_no_learning

max_L1        # maximum likelihood
pi_hat_1      # pi-hat: the parameter value where likelihood is maximized
LR1           # likelihood ratio vs. pi = 0.5 (no learning)

# Experiment 2 (7 successes in 10 trials)
max_L2 <- max(L.exp2)
# Find the \hat_pi => the max likelihood
pi_hat_2 <- pi_candidates[which.max(L.exp2)]
L2_at_no_learning <- dbinom(7, 10, 0.5)
LR2 <- max_L2 / L2_at_no_learning

max_L2
pi_hat_2
LR2
```

*Evidence statement*

Question: For Experiment2 (7 success over 10 trials) make a statement about the evidence relative to the following criteria

If LR < 20 => little or no evidence of learning
If LR $\geq 20$ => some evidence of learning

**Model answer for the evidence statement:** with LR $\approx$ 2.28 (well below 20), the data from Experiment 2 provide little or no evidence of learning. A worm getting 7 correct out of 10 trials is not a rare enough result under the "no learning" ($\pi = 0.5$) model to be convincing on its own.

**Note on the maximum-likelihood estimate:** 

For a simple binomial count, $\hat{\pi} = x/n$ exactly, that's why $\hat{\pi}_1 = 9/10 = 0.9$ and $\hat{\pi}_2 = 7/10 = 0.7$ come out as the maximum of the candidate list above. 

*Take-home message*: for a binomial count, $\hat{\pi} = x/n$ — always, exactly, no search required. 

```{r ex4-evidence-statement}
# TAKE-HOME MESSAGE: for ANY binomial count (x successes in n trials)
# the maximum likelihood estimate of the success probability is always
# pi.hat = x / n, EXACTLY. This is a general mathematical fact about the
# binomial distribution. Whenever you have simple binomial count data,
# you never need a grid search or a calculator to find pi.hat: just divide
# successes by trials.

mle_check <- function(x, n) {
  # optimize() searches continuously over pi in (0,1) for the value that
  # maximizes the binomial likelihood -- no candidate list required.
  fit <- optimize(function(pi) dbinom(x, n, pi), interval = c(0, 1), maximum = TRUE)
  data.frame(x = x, n = n,
             x_over_n       = round(x / n, 4),
             pi_hat_by_search = round(fit$maximum, 4))
}

# Four (x, n) pairs that have nothing to do with Experiment 1 or 2:
results <- rbind(
  mle_check(x = 3,  n = 8),
  mle_check(x = 15, n = 40),
  mle_check(x = 1,  n = 3),
  mle_check(x = 50, n = 200)
)
results
```

## Exercise 5: the cumulative distribution function (cdf)

What is a CDF?

CDF stands for cumulative distribution function. The CDF complements the Probability Density Function and provides a cumulative view of the probabilities linked to a random variable. Unlike the smooth curve of the PDF, the CDF appears as a step function, jumping at specific values. It shows the probability that a random variable will be less than or equal to a given value.

TO DO:

Can a planarian worm learn to avoid endpoint D? In a maze with 2 levels (Left right) a randomly moving worm will arrive at each of the four endpoints (A, B, C, or D) with equal probability. 

*What proportion of the time will a randomly moving worm arrive at endpoint D?*

Write the answer below 

$\pi = ?$

*What is the chance of never arriving at D in 6 trials? f0*

*What is the chance of arriving exactly once at D in 6 trials? f1*

*What is the probability that an untrained worm will arrives at D once or never in 6 trials.*

To calculate the answer, we need to accumulate probabilities.
We need to add the probability of no arrivals (f0) and the probability of arriving once at D (f1).

```{r ex5-by-hand}
# write the code to compute f0

# write the code to compute f1 

# write the code to compute $Pr{x <= 1}$
```


```{r ex5-cdf-function}
# The cumulative distribution function does the summing for us:
# Find the function and the code for doing it

```

*Handout: build the full pdf-and-cdf table, x = 0..6.*

```{r ex5-full-table}
x <- 0:6
f.x <- dbinom(x, size = 6, prob = 0.25)   # pdf
F.x <- pbinom(x, size = 6, prob = 0.25)   # cdf

pdf_cdf_table <- data.frame(x = x, `f(x)` = round(f.x, 4), `F(x)` = round(F.x, 4),
                             check.names = FALSE)
pdf_cdf_table
```

```{r ex5-cdf-barplot, fig.cap="Binomial cdf, n=6, p=0.25"}
barplot(F.x, names.arg = x, xlab = "x", ylab = "F(x) = Pr{X <= x}",
        col = "darkorange3", main = "Cumulative distribution, n = 6, p = 0.25")
```

## Exercise 6: the upper tail, $Pr\{X > x\}$

In the previous example we looked at $Pr{X < x}$ the probability of X or less successes in n trials. Let's say we are interested in $Pr{X > x}$ the probability of more than X successes. To obtain this probability we calculate the cumulative probability $Pr{X < x}$, then subtract it from 1, to obtain Pr{X > x}. The CDF accumulates (sums) all individual probabilities from the lowest possible outcome up to $x$

$$
Pr{X > x} = 1 - \sum_{k = 0}^{x} f(x)
$$
Let's say x = 6 success, so possible outcomes are 0, 1, 2, 3, 4, 5, 6 successes.

$$
Pr{X > x} = 1 - f(0) - f(1) - f(2) - f(3) - f(4) - f(5) - f(6)
$$

```{r illustrate-cdf}
################## COPY AND RUN THE WHOLE CHUNK #################
# Parameters for the Binomial distribution
n <- 6           # Number of trials
p <- 0.25        # Probability of success
x_cutoff <- 2    # Threshold x (e.g., calculating P(X > 2))

# 1. Outcomes and Probability Mass Function (PMF)
x <- 0:n
f_x <- dbinom(x, size = n, prob = p)

# 2. Cumulative Probability F(x) and Upper Tail Probability 1 - F(x)
cdf_x <- pbinom(x_cutoff, size = n, prob = p)       # P(X <= x_cutoff)
upper_tail <- 1 - cdf_x                             # P(X > x_cutoff)

# 3. Assign colors: Blue for X <= x_cutoff, Orange for X > x_cutoff
bar_colors <- ifelse(x <= x_cutoff, "skyblue", "orange")

# 4. Generate the plot with expanded y-limits (+ 0.20 headroom)
bp <- barplot(
  f_x,
  names.arg = x,
  col = bar_colors,
  border = "black",
  xlab = "Number of Successes (k)",
  ylab = "Probability f(k)",
  main = bquote("Illustration of " ~ P(X > .(x_cutoff)) == 1 - F(.(x_cutoff))),
  ylim = c(0, max(f_x) + 0.20)  # Increased height to prevent legend overlap
)

# 5. Add exact probability labels on top of each bar
text(
  x = bp,
  y = f_x + 0.015,
  labels = round(f_x, 4),
  cex = 0.8
)

# 6. Add legend at topright (where right-tail bars are small) without border box
legend(
  "topright",
  legend = c(
    paste0("P(X <= ", x_cutoff, ") = F(", x_cutoff, ") = ", round(cdf_x, 4)),
    paste0("P(X > ", x_cutoff, ") = 1 - F(", x_cutoff, ") = ", round(upper_tail, 4))
  ),
  fill = c("skyblue", "orange"),
  bty = "n",  # Removes the solid box around legend
  cex = 0.85
)
# Print values to console
cat(sprintf("F(%d) = P(X <= %d) = %.5f\n", x_cutoff, x_cutoff, cdf_x))
cat(sprintf("1 - F(%d) = P(X > %d) = %.5f\n", x_cutoff, x_cutoff, upper_tail))
```

*Handout: Try this yourself. "Can a planarian worm arrive at endpoint A 5 or more times in 6 trials?*

Compute $Pr\{X > 4\}$ two ways: (1) by summing the pdf's upper tail, and via (2) $1 - F(4)$."*

```{r ex6-upper-tail}
# Write the code for probability of 5 success in 6 trials with probability by trial being prob = 0.25 (f5)
f5 <- dbinom(NA, size = NA, prob = NA)
# Write the code for probability of 6 success in 6 trials with probability by trial being prob = 0.25 (f6)

f6 <- dbinom(NA, size = NA, prob = NA)

f5
f6

# Method 1: sum the tail of the pdf directly
Pr_X_gt_4_pdf <- f5 + f6
Pr_X_gt_4_pdf

# Method 2: use the cdf, which is faster once n gets large
Pr_X_gt_4_cdf <- 1 - pbinom(NA, size = NA, prob = NA)
Pr_X_gt_4_cdf

# Confirm the two methods agree
all.equal(Pr_X_gt_4_pdf, Pr_X_gt_4_cdf)
```

## Exercise 7: a harder maze, and comparing experimental designs

*Handout:*

(1) Draw a 3 tier maze with 8 possible endpoints (A-H). success = arrival at H

(2) What is the expected arrive rate at H by chance on each run through the maze?

$p_H = NA$

(3) What is the chance of 4 correct arrivals at H in 4 trials? ($Pr\{X=4\}$)

(4) What is the chance of 3 or more correct endpoints in 4 trials? ($Pr\{X>2\}$)

(4) Compare this design against the 2-tier/6-trial design from Exercises 5-6.*

```{r ex7-three-tier-maze}
p_H <- NA          
# expected arrival rate at H by chance, one of 8 endpoints
p_H

# write the code for the chance of 4 correct arrivals at H in 4 trials? ($Pr\{X=4\}$)
Pr_X_eq_4 <- dbinom(NA, size = NA, prob = p_H)
Pr_X_eq_4

# write the code for the chance of 3 or more correct in 4 trials ($Pr\{X>2\}$)
Pr_X_gt_2 <- 1 - pbinom(NA, size = 4, prob = p_H)
Pr_X_gt_2
```

*Handout: "At what value of x does $Pr\{X>x\}$ fall below 5% in each maze? Which design is more efficient (fewer trials to reach the 5% criterion)?"*

```{r ex7-design-comparison}
# 2-tier maze: n = 6 trials, p = 0.25 (Exercises 5-6)
x2 <- 0:6

# Calculate the probability of getting strictly more than $x_2$ successes out of 6 trials 
upper_tail_2tier <- 1 - pbinom(x2, size = 6, prob = 0.25)

# Put the successes and the probabilities together in a dataframe
data.frame(x = x2, `Pr{X>x}` = round(upper_tail_2tier, 4), check.names = FALSE)

# Find the x success where the probability is below 5% (0.05)
x2[upper_tail_2tier < 0.05]
x2_crit <- min(x2[upper_tail_2tier < 0.05])
x2_crit

# 3-tier maze: n = 4 trials, p = 1/8
x3 <- 0:4
# Calculate the probability of getting strictly more than $x_3$ successes out of 4 trials
upper_tail_3tier <- 1 - pbinom(x3, size = 4, prob = p_H)

# Put the successes and the probabilities together in a dataframe
data.frame(x = x3, `Pr{X>x}` = round(upper_tail_3tier, 4), check.names = FALSE)

# Find the x success where the probability is below 5% (0.05)
x3[upper_tail_3tier < 0.05]
x3_crit <- min(x3[upper_tail_3tier < 0.05])
x3_crit   
# smallest x where the 3-tier maze crosses below 5%
```

**Model answer:** the 2-tier maze needs $x=?$ (out of 6 trials) to fall below the 5% criterion; the 3-tier maze needs only $x=?$ (out of 4 trials).

The n-tier maze is the more efficient design. It reaches statistical significance with fewer trials

## Exercise 8: the chi-squared distribution and the G-statistic

The $G$-statistic (also called the Likelihood Ratio Test statistic) is a single number that measures how much better your best explanation fits the data compared to pure random chance.

Imagine you are testing whether an animal in a T-maze is actually learning or just guessing randomly:

Null Model (Chance): Assumes no learning ($\pi = 0.5$).

Observation (data): observed success rate (e.g., $\pi = 0.7$).

The Likelihood Ratio ($LR$) tells you how many times more plausible the best-fit model is compared to pure chance.

Likelihood no learning = 0.2668279

Likelihood observed success rate = 0.1171875

Likelihood ratio (LR) = $\frac{0.2668279}{0.1171875} = 2.28$

*For this Experiment the data was about $2.28$ times more likely under*

Is the difference large enough?

To evaluate this we use the G statistic = $2 ln(LR)$.

The $G$-statistic converts that raw Likelihood Ratio into a standardized statistical scale

Compute $G = 2\ln(LR)$ and its chi-squared p-value.

```{r ex8-G-statistic}
# We are getting variables from Exercise 4 (exact, unrounded)
# Experiment 2 (7 successes in 10 trials)
# max_L2 is max(L.exp2)
# L2_at_no_learning <- dbinom(7, 10, 0.5)
LR_exp2 <- max_L2 / L2_at_no_learning
LR_exp2

G_exp2 <- 2 * log(LR_exp2)
G_exp2

# Upper-tail p-value from the chi-squared distribution, df = 1
p_exp2 <- 1 - pchisq(G_exp2, df = 1)
p_exp2
```

Because $G$ matches the $\chi^2$ curve, you can instantly look up a $p$-value to answer: "What is the probability of getting a $G$ score this high purely by random luck?"

$G \approx 0$: Your model is no better than guessing.

Large $G$: Your model fits significantly better than guessing.

*Handout trial run: "As a trial run, we'll use a value of G = 3.84."*

We calculate probabilities for the $\chi^2$ distribution, which is used for hypothesis testing. As an example, we’ll use the coin-tossing experiment that resulted in 7 successes in 10 trials.

```{r ex8-trial-run}
# We want the probability in the upper tail Pr{ \chi^2 (df) > G}.
# We want the probability in the upper tail Pr{ \chi^2 (1) > G}.
# we define G_trial
G_trial <- 3.84

# This is the LOWER tail, Pr{chisq(1) <= G} =~ 0.95
pchisq(G_trial, df = 1)         
# this is the LOWER tail, Pr{chisq(1) <= G} =~ 0.95

1 - pchisq(G_trial, df = 1)     
# this is the UPPER tail we actually want, ~0.05
```

*Handout: more G/df combinations to evaluate.*

```{r ex8-more-G-values}
G_values <- c(5.67, 10.23)
df_values <- c(3, 5)
p_values <- 1 - pchisq(G_values, df_values)

data.frame(G = G_values, df = df_values, p = round(p_values, 4))
```

*Handout: "For the second coin tossing experiment, $G_{Experiment2} = 2\ln(LR) = 2\ln(2.3)$, df = 1, p = ___."*

```{r ex8-experiment2-rounded}
# Using the handout's own rounded LR = 2.3, for comparison with the exact value above
G_rounded <- 2 * log(2.3)
p_rounded <- 1 - pchisq(G_rounded, df = 1)

G_rounded
p_rounded
```


## Exercise 9: comparing the chi-squared pdf and cdf

*Handout: given a short table of $\chi^2$ pdf values, fill in the cdf column and compare a hand-sketch against the computed graph.*

Let's compare the probability density function pdf and the cumulative distribution function cdf of the chisquare distribution having 1 degree of freedom.

```{r ex9-table}
chisq_x <- c(0.5, 1, 2, 4, 8)

pdf_vals <- dchisq(chisq_x, df = 1)
cdf_vals <- pchisq(chisq_x, df = 1)
upper_vals <- 1 - cdf_vals

data.frame(x = chisq_x,
           `pdf = f(x)` = round(pdf_vals, 4),
           `cdf = F(x)` = round(cdf_vals, 4),
           `1 - cdf` = round(upper_vals, 4),
           check.names = FALSE)
```


```{r ex9-plots, fig.width=8, fig.height=4, fig.cap="Chi-squared (df=1) pdf and cdf"}
xx <- seq(0.05, 10, length.out = 300)   
# A smooth curve for comparison with a hand sketch

par(mfrow = c(1, 2))
plot(xx, dchisq(xx, df = 1), type = "l", lwd = 2, col = "firebrick",
     xlab = expression(chi^2), ylab = "f(x)", main = "pdf")
points(chisq_x, pdf_vals, pch = 19)

plot(xx, pchisq(xx, df = 1), type = "l", lwd = 2, col = "navy",
     xlab = expression(chi^2), ylab = "F(x)", main = "cdf")
points(chisq_x, cdf_vals, pch = 19)
par(mfrow = c(1, 1))
```

**Discussion: Comment on your initial sketch of the cdf, compared to the graph from calculations** 

(1) the cdf climbs very fast over the first unit or two then flattens out slowly across a long right tail.

(2) What else?

## Exercise 10: the t-distribution

The t-distribution is used to calculate probabilities when comparing two means, two slopes, etc.

**Why this matters:** the t-distribution is what lets you ask "are these two small groups actually different, or could that difference just be chance?" 

Common question in biology, comparing treatment vs. control, site A vs. site B, before vs. after. 

$t_{obs}$ stands for the observed $t$-statistic (this is your test score from your experimental data)

$$
t_{obs} = \frac{\text{Observed Difference} - \text{Hypothesized Difference}}{\text{Standard Error}}
$$

(1) $t_{obs} = 0$: Your sample result exactly matches the null hypothesis (no effect, no difference).

(2) Large $\vert{}t_{obs}\vert{}$ (e.g., $> 2.0$): The difference between the samples is unlikely to be pure random noise.

Compute the p-value using $t_{obs}$

*Handout: two-tailed p-value for $t_{obs}=-2.179$, df=12; one-tailed p-value for $t_{obs}=1.96$, df=300.*

(1) Two-Tailed Test (Non-Directional): You want to know if a sample mean is significantly different from a baseline (either higher OR lower). Your significance threshold ($\alpha = 0.05$) is split equally between both tails of the distribution ($0.025$ in the left tail, $0.025$ in the right tail).

(2) One-Tailed Test (Directional): You specifically predict before collecting data that the value will go in one exact direction (e.g., higher, better, faster). Because you only care about one side, the entire significance threshold ($\alpha = 0.05$) is placed into that single tail.

```{r ex10-t-distribution}
lefttail <- pt(-2.179, df = 12)
righttail <- 1 - pt(2.179, df = 12)
twotail <- lefttail + righttail
twotail # two-tailed p-value, t = -2.179, df = 12

onetail <- 1 - pt(1.96, df = 300)
onetail    # one-tailed p-value, t = 1.96, df = 300

# Two small groups of seedling heights (cm), grown under two light conditions
low_light  <- c(15.2, 16.1, 14.8, 15.9, 16.4)
high_light <- c(17.3, 18.0, 16.9, 17.6, 18.2)

t.test(low_light, high_light)
# `t.test()` just computes $t_{obs}$ and looks up its p-value in one step.
```

## Exercise 11: the F-distribution

The F-distribution is used to calculate probabilities when comparing several means (comparing three fertilizer treatments, four sites, five diets)

*Handout: upper-tail p-values for three $F_{obs}$ values, numerator df = 8, denominator df = 23.*

```{r ex11-F-distribution}
F_obs <- c(4.56, 2.28, 1.23)
p_F <- 1 - pf(F_obs, df1 = 8, df2 = 23)

data.frame(F_obs = F_obs, p = round(p_F, 4))
```

*Note* `aov()` just does the sums-of-squares and the `pf()` lookup for you.

## Exercise 12: the Z-statistic (Normal distribution)

the Z-statistic is the t-distribution's simpler cousin. You use it when the population mean and standard deviation are already known from a large reference dataset.

*Handout: two-tailed probability for $Z = 1.96$, $\mu=0$, $\sigma=1$.*

```{r ex12-Z}
Z <- 1.96
p_Z_twotailed <- 2 * (1 - pnorm(Z, mean = 0, sd = 1))
p_Z_twotailed
```


## Exercise 13: critical values (the inverse cdf)

In R, the Cumulative Distribution Function (CDF) for the Chi-Square ($\chi^2$) distribution is `pchisq()`.

`pchisq(q, df)`

Calculates the cumulative probability from $0$ up to $q$ for a given degree of freedom df

*Handout: find the inverse chi-squared cdf, and check that it returns 3.84 for df=1 at the 95th percentile (i.e., the 5% upper-tail critical value).*

```{r ex13-inverse-chisq}
qchisq(0.95, df = 1)   # should return ~3.84, matching Exercise 8's trial run
```

**Note:** `qchisq` is the *inverse* of `pchisq`. `pchisq` goes from a statistic $t_{obs}$ to a probability (p-value), `qchisq` goes from a probability back to a statistic.

## Exercise 14: extreme alpha levels, and building your own critical-value table

*Handout: An extremely conservative morphologist wants Type I error set at $\alpha$ = 0.0001. Find the critical $\chi^2$ value (df=1) at this level.*

```{r ex14-extreme-chisq}
qchisq(1 - 0.0001, df = 1)
```

*Handout: Table 3.1 — critical t-values at unusual alpha levels (0.002 and 0.0002), for df = 1, 5, 10. The handout's own worked row (df = 1) is given as a check: 159.156 and 1590.*

```{r ex14-t-table}
df_vals <- c(1, 5, 10)
alpha_vals <- c(0.002, 0.0002)

# One-tailed critical values, matching the convention used for qchisq above
# (an upper-tail area equal to alpha, not alpha/2 -- this is what reproduces
# the handout's own df=1 check values of 159.156 and 1590)
t_table <- outer(df_vals, alpha_vals, function(d, a) qt(1 - a, df = d))
dimnames(t_table) <- list(paste0("df=", df_vals), paste0("alpha=", alpha_vals))
round(t_table, 3)
```

```{r session-info}
sessionInfo()
```
