---
title: "Lab 2: Equations"
subtitle: "BIOL/OCSC 4605 · 7220 --- Statistical Analysis in Biology and Environmental Science"
author: "Student Name (replace)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    code_folding: show
    theme: flatly
    css: ../assets/css/course.css
    self_contained: true
params:
  week: 2
  module: "Part 1: Quantitative Background"
  due_date: "YYYY-MM-DD"
---

<!--
  ============================================================================
  STANDARD LAB TEMPLATE — BIOL/OCSC 4605 · 7220
  ============================================================================
  Things students need to get directly from the PDF: Figure 1 estradiol
  values for the Tamarin example, and the a/b allometric coefficients in the
  Notonecta table. Anderson & May's (1979) Table 4 parameter values for the
  Helminth example live in the source paper.
  ============================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 6, fig.height = 4
)

# Packages used in this lab -- uncomment / extend as needed.
# library(tidyverse)

set.seed(4605) # reproducible randomization / bootstrap results across labs
```

## Learning objectives

> Translating and using formal models of biological phenomena, expressed in the form of equations.

By the end of this lab, you should be able to:

1. Translate a described biological phenomenon into an explicit equation, step by step, following the 6-step recipe in Table 2.1.
2. Define every symbol in an equation in words, state its units, and connect a parameter's estimated value back to real data (theory or a fitting procedure such as least-squares regression).
3. Recognize that the same 6-step process applies across very different biological systems — physiology (metabolic rate), morphology (allometry), and epidemiology (parasite infection).

**Module:** Part 1: Quantitative Background &nbsp;|&nbsp; **Week:** 2 &nbsp;|&nbsp; **Key deliverable:** LaTeX / math notation in R Markdown

> Location: CSF BIOL 2218B Computer Lab · Time: 3:00–6:00 PM


## Background

A mathematical model is a translation: a described biological phenomenon, rewritten as symbols that can be estimated from data and checked against it. Every model in this lab is built with the same six-step recipe (**Table 2.1**, reproduced in full below), which forces you to be explicit about which variables matter, what relationship you are claiming holds between them, and what the parameters of that relationship actually mean in biological terms.

This lab works through that recipe four times: once fully worked here as a demonstration (**Kleiber's Law**, relating metabolic rate to body mass), and three more times as your own group exercises (Lab 2a): a behavioural/physiological example (tamarin maternal behaviour), a morphological/allometric example (*Notonecta* head length), and an epidemiological example (a helminth infection model, Anderson & May 1979).

**Table 2.1: the 6-step recipe**

1. Write the equation in symbolic form (symbols only).
2. Define each symbol in words and state units (if any).
3. State in words the idea expressed by the equation.
4. Write out parameter estimates, then write the equation with those parameter estimates substituted in. Parameter values sometimes come from theory, but more often from an estimation procedure such as least-squares regression.
5. Show that the units of all terms in the equation are equal.
6. Substitute specific values of the variable quantities to calculate the quantity of interest from the equation.

## Guided walkthrough

### Worked demonstration: Kleiber's Law

$$
\dot{E} = \alpha M^{\beta}
$$

1. **Equation:** $\dot{E} = \alpha M^{\beta}$

2. **Symbols:** $\dot{E}$ = resting metabolic rate; $M$ = body mass; $\alpha$ = scaling parameter (units that convert mass to metabolic rate); $\beta$ = unitless exponent.

3. **Parameter estimates (passerine birds):** $\alpha = 129$ kcal day$^{-1}$ kg$^{-0.724}$, $\beta = 0.724$ → $\dot{E} = 129\,M^{0.724}$.

5. **Units check:** 

Kleiber's Law (the equation)

$$\dot{E} = \alpha M^{\beta}$$

Then,

$$\alpha = \frac{\dot{E}}{M^\beta}$$

Replacing values (including the units)

What is $\alpha$?

$\alpha$ equals 129 kilocalories per day (day$^{-1}$) per kilogram to the power of 0.724 (kg$^{-0.724}$)

$$
129\ \mathrm{kcal}\,\mathrm{day}^{-1}\,\mathrm{kg}^{-0.724}
=
\frac{129\ \mathrm{kcal}\, \mathrm{day}^{-1}} {\mathrm{kg}^{-0.724}}
$$

kg$^{-0.724}$ cancels from the numerator and denominatoir, leaving 129 kcal day$^{-1}$ on both sides.


6. **Calculate:**

With this equation, you should now be able to calculate the minimum food energy requirement (in kilocalories per day) of a 20 gram (0.02 kg) canary.

```{r kleiber-demo}
alpha <- 129
beta <- 0.724

metabolic_rate <- function(M) alpha * M^beta

# Example: a 0.03 kg passerine bird
metabolic_rate(0.03)
```

**Calculate:**

With this equation, you should now be able to calculate the minimum food energy requirement (in kilocalories per day) of a 20 gram (0.02 kg) canary.

```{r kleiber-key-passerine}
alpha <- 129
beta <- 0.724
metabolic_rate <- function(M) alpha * M^beta

metabolic_rate(0.02)  
# a 20 g passerine ->
metabolic_rate(0.10)  
# a 100 g passerine ->
```

Everyone should do at least two of these on their own calculator, to become familiar with the mechanics.

**Verbal statement: The expected metabolic rate of a 20 gram bird, based on Laseiwski and Dawson's equation, is 7.59 kcal per day.**


**Try the same with Laseiwski and Dawson (1967) estimated α and β for non-passerine birds**

```{r kleiber-key-nonpasserine}
alpha <- 78.3
beta <- 0.723
metabolic_rate <- function(M) alpha * M^beta

metabolic_rate(0.02)  
# a 20 g non-passerine ->
metabolic_rate(0.10)  
# a 100 g non-passerine ->
```


Question: 

Compare this equation to that for passerines. Which will have the greater metabolic rate, a passerine bird, or a non-passerine of the same mass?


Try writing Kleiber's Law for $α = 540$, then calculate the metabolic rate for the 20 gram sparrow.

```{r kleiber-key-sparrow1}
alpha <- 540
beta <- 0.724
metabolic_rate <- function(M) alpha * M^beta

metabolic_rate(0.02)  
# a 20 g passerine (a sparrow)
```


This calculation is X times higher than the rate for a resting passerine.

**Question: Why?**

**The unit of $\alpha$ was incorrect**

Try calculating the metabolic rate (in kilojoules per day) of a 20 gram passerine (a sparrow) and a 500 gram passerine (a crow).

```{r kleiber-key-sparrow2}
alpha <- 540 # KJ/day
beta <- 0.724
metabolic_rate <- function(M) alpha * M^beta

metabolic_rate(0.02)  
# a 20 g passerine (a sparrow)
metabolic_rate(0.5)
# a 500 g passerine (a crow)
```

### Step 1 — Tamarin maternal behaviour (Lab 2a, example 1)

In a study of 12 red-bellied tamarins, urinary estradiol was the most abundant urinary estrogen during late pregnancy. Infant survival divided the group into "good" and "poor" mothers: 5 of 6 good mothers had two surviving infants at one week (including two without prior sibling-care experience), while 10 of 13 infants of poor mothers had died by day 0. Figure 1 in the handout plots mean urinary-estradiol concentration across the last 35 days of gestation for good vs. poor mothers.

**Your task:** propose a variable relating estradiol level to mothering outcome, and carry it through all six steps of Table 2.1.

The specific numeric values in Figure 1 are embedded in an image in the PDF handout and must be read directly off the graph — transcribe your own readings before completing Step 4.

```{r step-1-tamarin}

```

### Step 2 — *Notonecta* allometry (Lab 2a, example 2)

This is "example 104" from Simpson et al. (1960). The general allometric form is $y = a \cdot x^{b}$, where the exponent $b$ is called the *coefficient of allometry* (Greek *allo* = "other", *metron* = "measure") — it describes how strongly a body part stays in fixed proportion to total length as length changes. The measured part here is "synthlipsis to vertex", a measure of head length, in the water bug *Notonecta* (the backswimmer).

**Your task:** write the allometric equation with $x$ = total body length and $y$ = head length, define $a$ and $b$ in words with units, and fit $a$ and $b$ from the source table.

The specific $a$/$b$ coefficients are printed in a table inside the PDF and were not extractable as plain text — transcribe them directly from the handout.

```{r step-2-notonecta}

```

### Step 3 — Helminth infection model (Lab 2a, example 3)

This example comes from Anderson & May (1979), a foundational paper in mathematical epidemiology and parasitology. The handout directs you to their Table 4, which gives an equilibrium expression for the level of helminth infection in a host population.

**Your task:** locate Anderson & May (1979) Table 4, write the equilibrium-infection equation symbolically, define each parameter (transmission, mortality, recovery terms, etc.) in words with units, and substitute their reported estimates.

Table 4's specific parameter values are in the source paper itself, not the lab PDF — track down the full reference before class (see Background above for where published equations for Lab 2b can also be found).

```{r step-3-helminth}


```


## Exercises

Work through the following and show your code and output for each.

1. Complete Table 2.1's six steps for the **Tamarin** example above (Step 1 code chunk), citing your transcribed Figure 1 estradiol values.
2. Complete Table 2.1's six steps for the **Notonecta** example above (Step 2 code chunk), citing your transcribed allometric coefficients.
3. Complete Table 2.1's six steps for the **Helminth** example above (Step 3 code chunk), citing Anderson & May (1979) Table 4.

**Lab 2b — your own literature example.** After completing the three exercises above, find another example in the published scientific literature: an equation with **at least two variables and two parameters** (parameters will usually be shown as parameter estimates, rather than bare symbols). State the full reference (authors, title, journal, volume, and page number), then provide a complete translation and calculation by completing Steps 1–6 of Table 2.1 for it. If you use an equation from a source not already linked from the course website, note this explicitly on your report — it counts for extra credit.

```{r step-lab2b}

```

```{r session-info}
sessionInfo()
```
