---
title: "Lab 4: Probability Values by Randomization"
subtitle: "BIOL/OCSC 4605 · 7220 --- Statistical Analysis in Biology and Environmental Science"
author: "Your Name"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: flatly
    self_contained: true
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
params:
  week: 4
  module: "Part 2: Quantifying Uncertainty"
---

<!--
  ============================================================================
  INSTRUCTOR WALKTHROUGH -- BIOL/OCSC 4605 · 7220, Lab 4 "Probability Values
  by Randomization"
  ============================================================================
  Source: the course's Lab 4 handout ("Laboratory #4. Probability Values
  by Randomization"), a 10-page worksheet that builds a randomization test
  step by step on the Daphnia data (Box 9.5, Sokal and Rohlf 1995).

  This file follow the order of the printed handout, and every blank in the handout
  is answered with real, runnable R code. All computations use base R only 
  (no packages to install).

  Knit to PDF or HTML from the dropdown next to the Knit button in RStudio.
  ============================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                       fig.width = 6, fig.height = 4, fig.align = "center")
```

**The one idea of the whole lab:** if there is really no difference between two groups, then the group labels are meaningless, and shuffling the data between the groups should produce differences about as big as the one we observed. If our observed difference is *typical* of the shuffled differences, chance alone explains it. If it is *extreme*, chance is a poor explanation.

## The recipe: Table 4.1 and the flow diagram

*Handout: "Look again at Table 4.1 and then draw a flow diagram (boxes and arrows) showing how to do a randomization test."*

Table 4.1 gives the six steps of every randomization test:

1. Compute a statistic (the observed outcome).
2. Make the null hypothesis true by randomizing the data.
3. Re-compute the statistic to obtain an outcome when the null hypothesis is true.
4. Repeat this many times.
5. Construct a frequency distribution of outcomes when the null hypothesis is true.
6. Compare the observed outcome to the distribution of outcomes, in order to calculate a probability value.

Here is one possible flow diagram for the handout's blank box, drawn in base R. Step 4 is a loop: steps 2 and 3 are repeated many times.

```{r flow-diagram, fig.width=6.5, fig.height=7.2, fig.cap="Flow diagram for a randomization test (Table 4.1)"}
par(mar = c(0.5, 0.5, 0.5, 0.5))
plot(NA, xlim = c(0, 10), ylim = c(0, 14), axes = FALSE, xlab = "", ylab = "", asp = NA)

labels <- c("1. Compute the statistic\n(observed outcome)",
            "2. Make the null hypothesis true:\nrandomize the data",
            "3. Re-compute the statistic\nfor the randomized data",
            "4. Repeat steps 2 and 3\nmany times",
            "5. Build the frequency distribution\nof outcomes under the null",
            "6. Compare the observed outcome\nto the distribution: p-value")
ytop <- c(13.5, 11.2, 8.9, 6.6, 4.3, 2.0)    # top edge of each box
h <- 1.7                                      # box height

for (i in 1:6) {
  fill <- if (i == 6) "#cfe8cf" else if (i %in% c(2, 3, 4)) "#fde9c9" else "#d6e4f5"
  rect(2, ytop[i] - h, 8, ytop[i], col = fill, border = "grey30", lwd = 1.5)
  text(5, ytop[i] - h / 2, labels[i], cex = 0.95)
}
for (i in 1:5) {   # arrows from each box down to the next
  arrows(5, ytop[i] - h, 5, ytop[i + 1], length = 0.12, lwd = 1.5)
}
# the loop: from the bottom of step 4 back up to step 2 (the repeat)
segments(8, ytop[4] - h / 2, 9.4, ytop[4] - h / 2, lwd = 1.5)
segments(9.4, ytop[4] - h / 2, 9.4, ytop[2] - h / 2, lwd = 1.5)
arrows(9.4, ytop[2] - h / 2, 8, ytop[2] - h / 2, length = 0.12, lwd = 1.5)
text(9.55, (ytop[4] + ytop[2] - h) / 2, "repeat", srt = 90, cex = 0.85)
```

## Part 1: the research question and the data

*Handout: "Does the average age at beginning of reproduction in one strain of* Daphnia longispina *differ from that of another strain?" The data are Box 9.5 of Sokal and Rohlf (1995).*

**How many rows of data do you see?** The data file has **7 rows**, one age (in days) per row for each of the two strains. (The file also has a block of descriptive text *below* the 7 rows, which is why the handout's import command says `nrows=7`.)

**Which package do we use for this lab?** R.

### Define the data

The handout shows two ways to get the data into R. The chunk after them types the 7 pairs in directly, so the rest of this document runs anywhere.

1. Download the data directly to your computer

```{r load-by-download, eval=FALSE}
# 1. Download the Daphnia data file from the course website to your desktop.
# 2. Open the data and look at the data.
filelocation1 <- "/Users/abdoufofana/Desktop/DaphniaAges.txt"
daphnia <- read.table(filelocation1, nrows = 7, sep = "\t", header = FALSE)
names(daphnia) <- c("Strain1", "Strain2")

# How many rows of data do you see? 
# Write the code to find the number of rows
nrow(daphnia)
```

2. Load your data directly from an online source

```{r load-from-online, eval=FALSE}
#------------------------ OTHER METHODS ---------------------#
# 1. Copy the Daphnia data file link from the course website

"https://raw.githubusercontent.com/DavidCSchneider/StatisticalScience/main/Data/Labs/DaphniaAges.txt"

# 2. Load the Daphnia data in R using the link from the course website
filelocation2 <- "https://raw.githubusercontent.com/DavidCSchneider/StatisticalScience/main/Data/Labs/DaphniaAges.txt"

daphnia <- read.table(filelocation1, nrows = 7, sep = "\t", header = FALSE)
names(daphnia) <- c("Strain1", "Strain2")
```

*Handout typo: the handout's first import line writes `sep=" "` where the space between the quotes is meant to be a tab, which is easy to lose when copying. Writing `"\t"` makes the tab explicit. Also note the handout uses the object name `Daphnia` in one place and `daphnia` in another: R is case-sensitive, so these are two different objects. This script uses `daphnia` throughout.*

3. Type the data directly in R

```{r type-data}
# Data was typed in R directely
# Age (days) at first reproduction, Daphnia longispina (Box 9.5, Sokal & Rohlf 1995)
daphnia <- data.frame(
  Strain1 = c(7.2, 7.1, 9.1, 7.2, 7.3, 7.2, 7.5),
  Strain2 = c(8.8, 7.5, 7.7, 7.6, 7.4, 6.7, 7.2)
)
daphnia        # look at the data
nrow(daphnia)  # how many rows? 7
str(daphnia)
```

## Part 2: the observed statistic, AvDiff

*Handout: "Having defined the data, we calculate the difference in means between Strain1 and Strain2."*

The statistic is the difference between the two sample means, which the handout calls **AvDiff**. The `$` operator picks one column out of a data frame (`Filename$Variablename`).

```{r observed-statistic}
k1 <- mean(daphnia$Strain1)   # mean age at first reproduction, Strain 1
k2 <- mean(daphnia$Strain2)   # mean age at first reproduction, Strain 2
k3 <- k1 - k2                 # AvDiff = mean(Y1) - mean(Y2)

k1
k2
k3
```

*Handout: "Report the difference (in days) to 3 significant figures; then re-write it in units of hours."*

```{r avdiff-units}
# Example of how to report the difference in days to n significant figures
# difference in days = 0.87675456
# significant figures = 2
signif(0.87675456,2)

# Write the code to Report the difference (in days) k3 to 3 significant figures
AvDiff_days  <- signif(NA,NA)     # 3 significant figures
# Make sure you convert k3 days to hours 
AvDiff_hours <- signif(NA,NA)     # 24 hours per day

AvDiff_days    # days
AvDiff_hours   # hours
```

**Answer:** 
AvDiff = n days 
AvDiff = n hours
Write a sentence stating the difference in the average age at beginning of reproduction between Strain 1 and Strain 2

## Part 3: one random difference (RanDiff)

*Handout: "Could this difference be merely a matter of chance? To find out we randomize the data, to compute the difference in time to reproduction due to chance variation."*

The recipe for one random difference is: stack the two strains into one pool of 14 ages; draw 7 values at random for a fake "Strain 1" (RV1) and 7 for a fake "Strain 2" (RV2); take the difference of their means. If the null hypothesis is true, the strain labels carry no information, so the pool is all we have.

```{r one-random-difference}
# Step 1: stack both strains into one pool of 14 values
Y_all <- c(daphnia$Strain1, daphnia$Strain2)
Y_all
length(Y_all)

# Step 2: draw 7 values at random, with replacement, for each fake strain
RV1 <- sample(Y_all, size = 7, replace = TRUE)
RV2 <- sample(Y_all, size = 7, replace = TRUE)   # same command again, new random draw
RV1
RV2

# Step 3: the random difference in means
mean(RV1)
mean(RV2)
RanDiff <- mean(RV1) - mean(RV2)
RanDiff

# Step 4: store it in a vector (a "column") for later
RDvector <- RanDiff
RDvector
```

*Your random difference will differ from the one printed here, and from your neighbour's, because each run draws a new random sample. Run the chunk again to see it change.*

# Run the chunk `one-random-difference` again to see the difference

### Repeating the batch by hand

*Handout: "In order to assign a probability to the observed difference we need to accumulate many values of RanDiff." Run the batch several times and watch the vector grow.*

```{r accumulate-by-hand}
RDvector <- c()       # start with an empty vector

# Each pass through the batch adds one random difference. We run it 5 times
for (run in 1:5) {
  RV1 <- sample(Y_all, 7, TRUE)
  RV2 <- sample(Y_all, 7, TRUE)
  RanDiff <- mean(RV1) - mean(RV2)
  RDvector <- c(RanDiff, RDvector)   # stack the new value onto RDvector
}
RDvector
```

This works but is tedious: a real randomization test needs thousands of runs. That is what a *macro* is for.

## Part 4: the macro, 100 runs in one command

*Handout: "Define a sequence of commands; place into a file (a macro); execute the file a specified number of times." In R we can combine the whole batch into a single command with `replicate()`.*

`replicate(n, expression)` evaluates the expression `n` times and collects the results into a vector. The expression here is "one random difference", exactly as built in Part 3.

```{r macro-100-runs}
# One command: 100 random differences
RDvector <- c(replicate(100,mean(sample(Y_all, 7, TRUE)) - mean(sample(Y_all, 7, TRUE))))

length(RDvector)       # N = 100 randomized differences
round(head(RDvector, 10), 3)   # the first ten of them (rounded for display)
```

## Part 5: display the distribution

*Handout: "Display the distribution": a well-labelled histogram of the 100 random differences.* The handout code first, then a version that also marks the observed difference.

```{r histogram-handout, fig.cap="Figure 1. Randomized difference in mean age of first reproduction in two strains of Daphnia longispina."}
hist(RDvector, breaks = 15,
     xlab = "randomized difference",
     main = "Figure 1. Randomized difference in mean age of first reproduction in two strains of Daphnia longispina.",
     cex.main = 0.8)
```

```{r histogram-marked, fig.cap="The same histogram with the observed difference and its mirror image marked."}
hist(RDvector, breaks = 15,
     xlab = "randomized difference in means (days)",
     ylab = "frequency (number of runs)",
     main = "Randomization distribution, 100 runs",
     sub = "dashed lines: observed difference, + and -",
     col = "grey85", border = "white")
abline(v = c(-abs(k3), abs(k3)), col = "firebrick", lwd = 2, lty = 2)
```

**Read the plot and make a statement on wether the difference is a matter of chance:**

## Part 6: from the histogram to a p-value

Once we have a distribution, we can use it to calculate the probability of AvDiff, our observed statistic (St = -0.04286). We will use a two-tailed test

*Handout: "We will use a two-tailed test to calculate the probability of AvDiff, our observed statistic (St = -0.04286). We count the number of values in both tails, the left tail N(X <= St) and the right tail N(X >= St). We add the tails to obtain the proportion in both tails of the distribution."*

Why two tails? At the outset we had no idea whether Strain 1 would reproduce earlier or later than Strain 2, so a difference as extreme in *either* direction counts against the null hypothesis.

```{r sort-and-count}
# Smallest to largest (rounded the values for display)
round(sort(RDvector), 3)
# count in from each end towards -0.043 / +0.043
length(RDvector)     # N
k3                   # the observed difference, AvDiff
```

*Handout: "Use your package to compute N, n_neg, n_pos."*

```{r tail-counts}
N <- length(RDvector)                 # N = number of randomized differences

summary(RDvector < (-abs(k3)))   # lower tail: TRUE = number of values below
summary(RDvector > (abs(k3)))    # upper tail

# Floating-point numbers can differ in the 15th decimal place, so a value that
# "equals" the observed difference may not compare as equal. A tiny tolerance
# makes "as extreme as or more extreme than" work properly.
tol <- 1e-8

RDvector <= -abs(k3)
RDvector <= -abs(k3) + tol
n_neg <- sum(RDvector <= -abs(k3) + tol)   # as negative as -|AvDiff|, or more
n_pos <- sum(RDvector >=  abs(k3) - tol)   # as positive as +|AvDiff|, or more

N
n_neg
n_pos
```

*Handout: "Now compute the p-value for the observed outcome."*

```{r p-value}
pct_neg <- n_neg / N                # % of outcomes at or below -0.043
pct_pos <- n_pos / N                # % of outcomes at or above +0.043
p_value <- (n_neg + n_pos) / N      # two-tailed randomization p-value

pct_neg
pct_pos
p_value
```

**Answer:** 
n_neg/N = , 
n_pos/N = , 
So the two-tailed p-value is ** ? **.

Your numbers will differ a little from run to run, but they should land around 0.9. This is the estimate of the Type I error (the p-value) under the null hypothesis that the true difference is zero: about nine times in ten, shuffling the data gives a difference at least this large. There is **no evidence** that the two strains differ in age at first reproduction.

### Check against the F-distribution

Using the F-distribution to test whether the two strains differ in age at first reproduction

*Null hypothesis: the two strains have the same age at first reproduction*

*Handout: "It turns out that assumption free p-values by randomization in this example are close to the value (p = 0.908) calculated from the F-distribution with 1 and 12 degrees of freedom (F = 0.014)."*

```{r f-check}
pf(0.014, 1, 12, lower.tail = FALSE)  
# 1 is df1, the number of groups - 1
# 12 is the df2, total observation - 2
# the handout's command
# 0.014 is the F-ratio. This will be explained later

# pf() calculate a p-value to test wether the 2 strain 

# The same test done exactly (a two-sample t-test with equal variances; F = t^2):
t.test(daphnia$Strain1, daphnia$Strain2, var.equal = TRUE)
```


The calculation of this F-ratio (F1,12 = 0.014) will be explained later in the course, as will the assumptions for using the cdf.

```{r two-sample-t-test}
# The same test done exactly (a two-sample t-test with equal variances; F = t^2):
# Look at the data by strain first 
daphnia$Strain1
daphnia$Strain2

# T-test on the 2 samples
t.test(daphnia$Strain1, daphnia$Strain2, var.equal = TRUE)
# Find the p-value and compare with the F and Randomization approach
```

The randomization p-value (about 0.9) and the F-distribution p-value (0.908) agree closely, even though the randomization made no assumption about normal or homogeneous errors.

## The p-value is itself random (so how many runs are enough?)

We only did 100 runs "so that we can see what we are doing". The cost of that is that the p-value changes from run to run. Repeat the whole 100-run test eight times, then eight times with 10,000 runs each:

```{r p-value-noise}
one_test <- function(nruns) {
  rd <- replicate(nruns, mean(sample(Y_all, 7, TRUE)) - mean(sample(Y_all, 7, TRUE)))
  mean(abs(rd) >= abs(k3) - tol) 
  # two-tailed p-value 
  # mean(abs(rd) >= abs(k3) - tol) is simply counting mean differences <= -k3 and >= k3 (we did this in the previous session). Then calculate the frequency
  # n_neg <- sum(rd <= -abs(k3) + tol)
  # n_pos <- sum(rd >= abs(k3) - tol)
  # (n_neg + n_pos) / N	is exactly mean(abs(rd) >= abs(k3) - tol) = p-value
}

round(replicate(8, one_test(100)), 2)      # 100 runs
round(replicate(8, one_test(10000)), 3)    # 10,000 runs
```

With 100 runs the p-value moves around by a few hundredths; with 10,000 it barely moves. The conclusion (p is large) is the same, but if a p-value sat close to 0.05 the difference would matter.


## Write-up, question 2: guinea-pig litter size

*Handout: "Carry out a randomization test for data in Box 13.12 of Sokal and Rohlf (1995), which shows mean litter size of two strains of guinea pig, compared over 9 years. Assign a symbol to your statistic (difference of means). State the observed value of your statistic. Report your assumption-free p-value via randomization. Make a frequency distribution of at least 200 randomized outcomes, and label both axes."*

The data (average litter size in strains B and 13, from Sewall Wright's colony, 1916-1924) are in the course file `LitterSize.txt`

```{r litter-import}
# Load the LitterSize.txt file in R
# 1. Find the online location or computer location of the file. filelocation3 <- "location of the file" 
#filelocation3 <- ""

filelocation3 <- "https://raw.githubusercontent.com/DavidCSchneider/StatisticalScience/main/Data/Labs/DaphniaAges.txt"

litter <- read.table(filelocation3, nrows = 7, header = FALSE)
names(litter) <- c("StrainB", "Strain13")

# The correct code is below, because litter has 3 columns and the 3rd column is Year
# 2. Find the correct nrows = 
# 3. names(litter) <-  c("StrainB", "Strain13", "Year")

# 4. Look at the data by strain
litter$StrainB
litter$Strain13
```

**Step 1, the statistic.** Use the same statistic as the Daphnia example, the difference of means: $\Delta\bar{Y} = \bar{Y}_B - \bar{Y}_{13}$.

```{r litter-observed-stratistics}
# Use litter$StrainB and litter$Strain13 to calculate the mean by group and mean difference
k1 <- mean(daphnia$Strain1)   # Average litter size in guinea pigs, Strain B
k2 <- mean(daphnia$Strain2)   # Average litter size in guinea pigs, Strain 13
k3 <- k1 - k2                 # AvDiff = mean(Y1) - mean(Y2)

k1
k2
k3
```

**Steps 2 to 5, the randomization.** The pool has 18 values; each fake group gets 9 (the number of years). We use 200 runs.

```{r litter-randomization}
# 1. Pool the 18 values together
Y_litter <- c(litter$StrainB, litter$Strain13)   # pool of 18 values
length(Y_litter)

# 2. Change the 7 to the correct number of observation for litter
# mean(sample(Y_litter, , TRUE)

# 3. Change the number of replicates to 200 
# Make 200 randomized outcomes
RD_litter <- replicate(100,
                       mean(sample(Y_litter, 7, TRUE)) -
                       mean(sample(Y_litter, 7, TRUE)))
length(RD_litter)
```

```{r litter-histogram, fig.cap="Randomized difference in mean litter size of guinea-pig strains B and 13 (200 runs)."}
hist(RD_litter, breaks = 20,
     xlab = "Your x-axis label",
     ylab = "frequency (number of runs)",
     main = "Give a Title to your plot",
     sub = "dashed lines: observed difference, + and -",
     col = "grey85", border = "white")
abline(v = c(-abs(k3), abs(k3)), col = "firebrick", lwd = 2, lty = 2)
```

**Step 6, the p-value.**

```{r litter-p-value}
n_neg_L <- sum(RD_litter <= -abs(k3) + tol)
n_pos_L <- sum(RD_litter >=  abs(k3) - tol)
p_litter <- (n_neg_L + n_pos_L) / length(RD_litter)

n_neg_L
n_pos_L
p_litter
```

**Answer for the handout blanks:**

- Symbol for the statistic: $\Delta\bar{Y} = \bar{Y}_B - \bar{Y}_{13}$.

- Observed value: $\Delta\bar{Y}$ = NA

- Report your assumption-free p-value via randomization


```{r litter-save-pdf, eval=FALSE}
pdf("Lab4_LitterSize_randomization.pdf", width = 7, height = 5)
hist(RD_litter, breaks = 20,
     xlab = "randomized difference in mean litter size (strain B - strain 13)",
     ylab = "frequency (number of runs)",
     main = "Randomization distribution, 1,000 runs")
abline(v = c(-abs(k3), abs(k3)), col = "red", lwd = 2, lty = 2)
dev.off()
```

## Wrap-up

- [ ] Flow diagram (Table 4.1) discussed *before* any code was run
- [ ] Ran the one-random-difference chunk several times and saw it change
- [ ] 100-run macro run live; histogram displayed with the observed difference marked
- [ ] Count `n_neg`, `n_pos`, N and compute p-value
- [ ] Guinea-pig test: statistic symbol, observed value, p-value, labelled histogram of at least 200 runs

```{r session-info}
sessionInfo()
```
