---
title: "Lab 10: Logistic Regression (GLM)"
subtitle: "BIOL/OCSC 4605 · 7220 --- Statistical Analysis in Biology and Environmental Science"
author: "Student Name (replace)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    code_folding: show
    theme: flatly
    css: ../assets/css/course.css
    self_contained: true
params:
  week: 12
  module: "Part 5: The Generalized Linear Model"
  due_date: "YYYY-MM-DD"
---

<!--
  ============================================================================
  STANDARD LAB TEMPLATE — BIOL/OCSC 4605 · 7220
  ============================================================================
  This file was generated from templates/lab_template.Rmd. Keep the section
  structure and YAML options consistent across labs so students always know
  where to look. Replace every "(edit me)" / placeholder comment before
  distributing this lab.

  To regenerate ALL 13 labs from an updated version of this template, edit
  templates/lab_template.Rmd and re-run the generator script described in the
  repo README, then re-apply any lab-specific edits you had made.
  ============================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 6, fig.height = 4
)

# Packages used in this lab -- uncomment / extend as needed.
# library(tidyverse)
# library(emmeans)
# library(car)
# library(boot)

set.seed(4605) # reproducible randomization / bootstrap results across labs
```

## Learning objectives

> Non-Gaussian response variables. Modeling binary/proportional biological outcomes using logit link functions.

By the end of this lab, you should be able to:

1. *(objective 1 --- edit me)*
2. *(objective 2 --- edit me)*
3. *(objective 3 --- edit me)*

**Module:** Part 5: The Generalized Linear Model &nbsp;|&nbsp; **Week:** 12 &nbsp;|&nbsp; **Key deliverable:** Fitted logistic model with interpretation

> Location: CSF BIOL 2218B Computer Lab


## Background

*(One to two paragraphs of conceptual background specific to this lab. Link
back to the relevant section of Schneider's*
[*Statistical Science*](https://davidcschneider.github.io/StatisticalScience/)*.)*

## Data

*(Describe or load the dataset used this week. Replace the placeholder code
below with the actual read-in step, or remove it if this lab has no dataset,
e.g. Lab 1's physical-cards exercise.)*

```{r load-data}
# df <- read.csv(here::here("data", "FILENAME.csv"))
# str(df)
```

## Guided walkthrough

### Step 1 --- *(edit me)*

```{r step-1}

```

### Step 2 --- *(edit me)*

```{r step-2}

```

### Step 3 --- *(edit me)*

```{r step-3}

```

## Diagnostics / verification

*(Residual plots, randomization overlays, homogeneity checks, or other
verification appropriate to this lab's method. Primary tools this week:*
`glm(family = binomial)`*.)*

```{r diagnostics}

```

## Exercises

Work through the following and show your code and output for each.

1. *(exercise 1 --- edit me)*
2. *(exercise 2 --- edit me)*
3. *(exercise 3 --- edit me)*

## Submission questions

Answer in complete sentences below each question. These responses, together
with your code above, are graded against this lab's key deliverable:
**Fitted logistic model with interpretation**.

**Q1.** *(question --- edit me)*

> *(your answer here)*

**Q2.** *(question --- edit me)*

> *(your answer here)*

**Q3.** *(question --- edit me)*

> *(your answer here)*

## Submission checklist

- [ ] Code chunks all run top-to-bottom without error (Knit ▸ Knit to HTML)
- [ ] All placeholder text above has been replaced
- [ ] Submission questions answered in full sentences
- [ ] File renamed / submitted per course LMS instructions

---

```{r session-info}
sessionInfo()
```
